### 2016-02-09 Manuscript/BiologyLetters/response.tex

\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}

\setlength\parindent{0pt}

\begin{document}

\textbf{RE: Decision on Manuscript ID RSBL-2015-1003}\\
\bigskip
Dear Surayya Johar,\\
\bigskip
We are very grateful to the four referees for their helpful and constructive comments.
We have taken all of their comments on board, and respond to their points below. To improve the clarity of this document, we dealt with each comment in the order they appeared in your ``Decision on Manuscript'' email and we have included the reviewers comments in blue.

%TG: plus response to editors points:
% *Please also confirm in your cover letter whether all the figures are your own or whether permission has been obtained for their use
% *Please confirm whether your manuscript should be available via open access, as we note that this may be a requirement of your funder.
We would like to state that all the figures are ore our own and the the manuscript should be available via open access.
% *If you have any images that can be used to promote your article on social media (should this be accepted) please upload them as a supplementary file
Additionally, we attached a supplementary file named \texttt{article\_promotion.eps} for promoting our article (should it be accepted) on social media.

% *Please upload your original figure files as eps, tiff or jpeg files, rather than PDFs
%TG: TO DO (easy)

@@

As shown in ESM1 Figure 1, the number of OTUs with morphological cladistic data extracted from the Google Scholar searches seems to plateau really quickly (i.e. before the first 100 matches).
Additionally, as Graeme Lloyd mentions, ``older studies tend to be smaller (due to computing limitations) and to commonly reoccur anyway due to data set reuse'' (Reviewer 4, comment 1).
Following these two observations, we think that including the 20 most recent results % NC: Is that what you did? TG: yep.
 for each order (and higher taxonomic level) was a good balance between the time spent looking for papers, the number of OTUs extracted and the repeatability of the study.
We are aware however that we might have missed some specific studies (see Reviewer 2 comment 2) but are confident that our protocol gives a good overall picture of the data availability and that the three repositories used before the Google Scholar search (MorphoBank, Graeme Lloyd's and Ross Mounce's) contained most of the available data.
In fact, as mentioned in ESM 1, part 1 ``among the 660 papers [collected from the Google Scholar search], only 50 contained a total of 425 extra living OTUs''.

\item{\textcolor{blue}{Similarly, I find the 2010 date cutoff troubling, because I'm not sure that the argument that these recent studies contain a fraction of the previously published characters and OTUs allows you to make a comprehensive statement about the availability of character data. What about the fraction that isn't included? This is a question that needs to be better addressed before you will convince readers that this is really all the data that is out there.}}

We chose this cutoff date to filter our analysis towards accessible data only (see Reviewer 2 comment 2).
Additionally, we observed that in newer studies (i.e. post 2010), living taxa with anatomical data were rarely added and usually reused from former studies (with sometimes the addition of extra characters and extra living taxa).
We modified the sentence in the ESM1 part 1:

@@

\item{\textcolor{blue}{P3, L78: ``since'' should be ``because.'' Since should be used for time, not causation.}}
We replaced ``since'' with ``because'' throughout the manuscript as appropriate.

\item{\textcolor{blue}{P3, L85: ``not-applicable'' should not be hyphenated.}}
We removed the hyphen.

\item{\textcolor{blue}{P3, L120: How much of a difference does the 25\% or less coverage make to the capacity of Total Evidence phylogenetic analyses to resolve relationships? Does your simulation study have a way of quantifying or describing the probability of getting the ``right'' answer with this little data? It would also help if you could show in your data table what the clades were that had better than 75\% coverage. It's not immediately evident from reading Table 1.}}

The straight relationship between the living species coverage and the Total Evidence phylogeny topology is hard to quantify precisely and depends on both the metric used for measuring topological differences and the method used to infer the tree.
Our simulation showed that topology tend to get closer to a random one as morphological data is missing but that in the case of the missing living species, the other parameters we tested (missing data in the fossil record and number of morphological characters) had no significant effect (i.e. when a certain amount of living taxa are missing, the other parameters do not have any extra influence).
In our simulations we tested 5 different level of missing living taxa (0\%, 10\%, 25\%, 50\% and 75\% of missing living taxa) and we observed that the three first levels had no significant negative effect on topology in relation with the amount of missing data in the fossil record and the number of morphological characters but that the fifth level (75\% of missing living taxa, i.e. 25\% coverage) had a strong negative effect.

@@

We removed the analysis with a threshold of OTUs with at least 100 characters and replaced it with the analysis without any minimal number of morphological characters.
We retracted the paragraph about the character threshold analysis and updated the results of figure 1 and table 1 to contain all living OTUs; we removed the redundant ESM 2 and we updated the figures

\item{\textcolor{blue}{Along these lines, a 100 character limit may seem rational but this is actually, from a practical perspective, very large. molecular matrices of this size would, of course, be laughed at but many of the most robust phylogenetic studies I'm familiar with come in well under this size. I suspect, but cannot be sure, that this partially explains the poor species level coverage the authors find; higher level studies can readily accumulate large numbers of characters because vastly different morphologies are being compared, often using families or genera as exemplars, leading to an abundance of easily defined characters. Species level studies, on the other, are often focused within families or genera to make them more manageable, but here characters are quickly exhausted. To give some random examples that I'm familiar with, Salles (1992) species-level study of Felidae (all species) was based on 44 characters, Wang and Carranza- Casta\~{n}eda's 2008 study of Mephitidae (\textasciitilde50\% of species but all genera) was based on 38 characters, and Bryant et al's 1993 study of Mustelid genera was of 46 characters. These are all important studies, but do not feature here because of the 100 character limit. At the very least I think this results in a downplaying of previous phylogenetic efforts by morphologists in the results. At worst it misrepresents the state of affairs dramatically. Is 100 really the minimum number you want to go with at all phylogenetic scales?}}

By removing the characters threshold (see above) we now included the taxa coded in Wang and Carranza-Casta\~{n}eda 2008 (matrix reference GS2014-WCC) and other matrices ignored in the former analysis increasing the number of Carnivora from 42 species (14\% of total number of species) to 76 (27\%).
However, the two other matrices mentioned by reviewer 2 (Salles 1992 and Bryant et al. 1993) where not included in the present analysis.
This is due to two main reasons: firstly, these two matrices where not available on any of the matrices repositories (either Graeme Lloyd's, Ross Mounce's or MorphoBank's) and (2) secondly, they didn't matched with our additional Google Scholars search parameters (i.e. matrices published between 2010 and 2015 with accessible data).
Following this reviewer's comment as well as reviewer 4's we added a paragraph in the discussion to clearly state that we are aware of data sampling biases:

@@

We totally agree with this comment.
In fact, among the 286 matrices used in this analysis, we found 5228 unique OTUs including 3627 fossil ones which is pretty impressive knowing the amount of work required to collect data for a single specimen!
However, in this study, we focused only on the amount of living taxa (1601 unique OTUs containing 847 species) as a way to solve the problems associated with the lack of living taxa on tree topology described in Guillerme \& Cooper 2016.

\item{\textcolor{blue}{Finally, I would add: who the heck submits a paper for a special volume 6 weeks early?!?!? You make the rest of us look bad!}}
This paper formed part of my PhD thesis that I submitted in September which partially explains the swift submission!
Sorry for making the rest of the authors submitting to the issue look bad.

\end{enumerate}

\subsection{Reviewer 3 specific comments:}
\begin{enumerate}
\item{\textcolor{blue}{Lines 25-26: ``We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies.''
In a way, this is more than ``total evidence''.
The initial total evidence paradigm was simply to use every character you could (molecular, morphological, even behavioral).
We now are talking about tip-dating and fossilized birth-death (FBD) approaches (e.g., Heath et al. 2014 PNAS 111:E2957) that want to use all of the available data.
After all, one can do a total evidence analysis with just node-dating techniques rather than with tip-dating, and one can do it without worrying about branch length priors.
However: if we want to use all of the data to calibrate branch lengths for all taxa (fossil and extant), then we need ``total evidence'' to do tip-dating and FBD properly).
As tip-dating is the main theme of this issue, that should be linked to the total evidence explicitly here and elsewhere in the manuscript.}}

We are aware of the difference between the ``Total Evidence method'' (i.e. a way to combine data in a matrix like supermatrix) and ``tip-dating method'' (i.e. a phylogenetic dating method, like node dating).
Both methods can (and are often) be used together as in Ronquist et al 2012 (Systematics Biology) but we are aware that they can also be used really well separately.
However, this study focuses \textit{indeed} on the ``Total Evidence method'' (i.e. a way to combine molecular and discrete anatomical data together) and not on the ``tip-dating'' one because it attempts to quantify the sate of the problem of missing data in living taxa underlined in Guillerme \& Cooper 2016 (again, a paper focusing on the ``Total Evidence method'', not including ``tip-dating'' nor ``node datin'' analysis).
Nonetheless, we do understand this reviewer's point and added the mention of ``Total Evidence tip-dated phylogenies'' which is undoubtedly a great avenue of application of both methods.

We modified the sentence in the abstract as follow: ``We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence tip-dated phylogenies.'' p@ lines@@

And clarified the third sentence of the introduction as follows:
``One increasingly popular method, the Total Evidence method [3], combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method (e.g. [4, 3, 5, 1, 6]), producing a chronogram with living and fossil taxa at the tips.'' p@ lines@@

@@

We removed the character threshold in the whole analysis and specified what we meant by ``data accessibility'' in the discussion (see Reviewer 2 comment 2).

\item{\textcolor{blue}{Reviewer 3 (Pete Wagner)'s major problem - with which I agree - is semantic, and thus easily fixable with a few explicit caveats (e.g., ``by cladistic we mean...'').}}

We modified the mention of ``cladistic characters'' to ``anatomical characters'' throughout the manuscript (see reviewer 3 comment 1).

@@

\item{\textcolor{blue}{ESM 2 - ``The following section contains supplementary results to the main body: the available data structure using the NTI and the PD metric'' Abbreviations should be introduced when first used, not later as they are here.}}
We changed this to ``using the Net Relatedness Index (NRI) and the Nearest Taxon Index (NTI) metrics''.

\end{enumerate}

We hope we have responded to all these comments appropriately. Please let us know if you require any further information,\\
\bigskip

% Editorial office comments to authors:
% Please ensure that you include; 
% *An author contributions section which fulfils all *four* of our criteria given here https://royalsociety.org/journals/ethics-policies/openness/, including a sentence stating all authors agree to be held accountable for the content therein and approve the final version of the manuscript 
%TG: Ok, we I'm not sure here. Is it just an automatic message: did all the author contributed well enough to be author (yes!); or is it that we have to modify the author contribution list to the exact wordings they put in this URL (seems odd though).

% *A funding statement which is the same as that given in ScholarOne
%TG: Yes, is matching exactly.

% *Please also confirm in your cover letter whether all the figures are your own or whether permission has been obtained for their use
%TG: TO DO (yes)
% *Please upload your original figure files as eps, tiff or jpeg files, rather than PDFs
%TG: TO DO (easy)
% *If you have any images that can be used to promote your article on social media (should this be accepted) please upload them as a supplementary file
%TG: what about the "phylogeny" with the duck and the pig?
% *Please confirm whether your manuscript should be available via open access, as we note that this may be a requirement of your funder.
%TG: TO DO (yes)

### 2016-02-09 Manuscript/BiologyLetters/Missing_mammals-draft.tex
Response wrapped up

We investigated the amount of available anatomical characters for living mammals and how this data was phylogenetically distributed across orders.
22 of 28 mammalian orders have \textless 25\% species with available anatomical characters; this has implications for the accurate placement of fossil taxa, although the issue is less pronounced at higher taxonomic levels. 
In most orders, species with available data are randomly distributed across the phylogeny, which may reduce the impact of the problem.
We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies. 
\end{abstract}

@@

There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data %TG: I left "morphological data" here since it TEM it doesn't have to be only anatomical characters
 from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a phylogeny with living and fossil taxa at the tips. 
A downside of this method is that it requires molecular data for living taxa and morphological data for both living and fossil taxa. %TG: same comment as above
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may restrict the amount of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
This is because (1) fossils will not be placed accurately within the correct clade if it contains no anatomical characters for living taxa; and (2) fossils have a higher probability of branching within clades with more anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

>>>>

We investigated the amount of available anatomical characters for living mammals and how this data was phylogenetically distributed across orders.
22 of 28 mammalian orders have \textless 25\% species with available anatomical characters; this has implications for the accurate placement of fossil taxa, although the issue is less pronounced at higher taxonomic levels. 
In most orders, species with available data are randomly distributed across the phylogeny, which may reduce the impact of the problem.
We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence tip-dated phylogenies. 
\end{abstract}

@@

There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data %TG: I left "morphological data" here since it TEM it doesn't have to be only anatomical characters
 from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method  (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a chronogram with living and fossil taxa at the tips. 
A downside of this method is that it requires molecular data for living taxa and morphological data for both living and fossil taxa. %TG: same comment as above
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may restrict the amount of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
This is because (1) fossils will not be placed accurately within the correct clade if it contains no anatomical characters for living taxa; and (2) fossils have a higher probability of branching within clades with more anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.
